---
title: 'Reproducible Research : Peer Assessment 1'
author: "Rajesh Pyakurel"
date: "4th June 2016"
output: html_document
---

## Loading and preprocessing the data

Loading Necessary Graphics Package

```{r}
library (ggplot2)
library (lattice)
```

#### 1. Load the data (i.e. read.csv()) 
```{r}
#### The data is downloaded from source provided in Coursera Project 1 that link to
#### https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

activity <- read.csv("./data/activity.csv",  sep = ",", header = TRUE, stringsAsFactors = FALSE)
head(activity)
tail(activity)
```

#### 2. Process/transform the data suitable for analysis
Looking into head and tail of activity, it is not necessary to transform the data. Wherever required "na.rm=T" can be used.

## What is mean total number of steps taken per day?

#### 1. Calculate the total number of steps taken per day
```{r echo = TRUE}
#### aggregate function is used to calculate sum of steps according to date.
step_activity<-aggregate(steps~date,activity,sum,na.rm=T)
```

#### 2. Creating a histogram of the total number of steps taken each day
```{r histogram,echo = TRUE, fig.height=6}
hist(step_activity$steps,breaks=12, col="red", xlab = "Total number of steps taken each day", main = "Histogram of total steps per day")
```

#### 3. Calculate and report the mean and median of the total number of steps taken per day
```{r echo = TRUE}
mean(step_activity$steps)
median(step_activity$steps)
summary(step_activity$steps)
```

##  What is the average daily activity pattern?

#### 1. Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
``` {r Time_Series_Plot, fig.height=6}
#### aggregate function is used to calculate mean of steps according to interval. Our interval is of 5 steps
step_interval_activity<- aggregate(steps ~ interval, activity, mean,na.rm=T)

g <- ggplot(step_interval_activity, aes(interval, steps))
g + geom_line(size=1,col="red")+labs(title="Average steps for each interval")
```

#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
step_interval_activity[step_interval_activity$steps == max(step_interval_activity$steps), ]

```

## Imputing missing values  

#### 1. The total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
numMissingValues <- length(which(is.na(activity$steps)))
numMissingValues 
```
The total number of rows with NAs is `r numMissingValues`.

#### 2. Devise a strategy for filling in all of the missing values in the dataset. 

#### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

The dataset is created to fill all missing values as follows

```{r}
### I am using R version 3.2.2.So Impute package can not be used.
### So I assign the missing values as follows

new_activity<-activity  # new dataset without NAs
interval<-tapply(activity$steps, activity$interval, mean, na.rm=T)
for (i in 1:nrow(activity)){
  if(is.na(activity$steps[i])){
     new_activity$steps[i]<- interval[[as.character(activity[i, "interval"])]]
  }
}
#### Checking the number of missing value in new dataset
numMissingValues <- length(which(is.na(new_activity$steps)))
numMissingValues
```
After imputing, the count of NAs values is `r numMissingValues`.

#### 4. Make a histogram of the total number of steps taken each day 
```{r histogram_No_missing_data, fig.height=6}

step_activity_imputed<-tapply(new_activity$steps, new_activity$date, sum)
hist(step_activity_imputed,breaks=16,col = "red" ,xlab = "Total number of steps taken each day", main = "Histogram of total steps per day with imputation")


```

#### 5. Calculate and report the mean and median total number of steps taken per day
```{r}
mean(step_activity_imputed)
median(step_activity_imputed)
summary(step_activity_imputed)
```
There isnont significant change after populating N/A values. The median becomes little higher than before but summary seems to be same.

## Are there differences in activity patterns between weekdays and weekends?

#### 1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" 
```{r}
new_activity$dateType <-  ifelse(as.POSIXlt(new_activity$date)$wday %in% c(0,6), 'weekend', 'weekday')
table(new_activity$dateType)
```

#### 2. Panel plot for a time series plot of the 5-minute interval (x-axis) and the average number of steps  taken, averaged across all weekday days or weekend days (y-axis)
```{r panel_plot_weekdays_vs_weekend,fig.height=6}
average_step_interval_new_activity <- aggregate(steps ~ interval + dateType, data=new_activity, mean)
average_step_interval_new_activity <- aggregate(new_activity$steps, list(interval = as.numeric(as.character(new_activity$interval)), weekdays=new_activity$dateType),FUN = "mean")
names(average_step_interval_new_activity)[3] <- "meanOfSteps"
xyplot(average_step_interval_new_activity$meanOfSteps ~ average_step_interval_new_activity$interval | average_step_interval_new_activity$weekdays, 
       layout = c(1, 2), type = "l", 
       xlab = "Interval", ylab = "Number of steps")
```